---
title: "1. Selecting CpGs for Age Prediction"
author: "Ellie Weise"
date: "2024-09-20"
output: html_document
---

```{r,echo=F,warning=F,message=F}
##load data
load("Input/allCpGs_AddAge1629Samples.rda")
sexdata <- read.csv("Input/MethylWild_Sex_AgeData.csv")

#load libraries
library(tidyverse)
library(glmnet)
library(MASS)
library(wesanderson)
```


There are many sources of methylation variation among individuals in a population, from environmental effects to stress and life history. However, a small number of methylation sites (CpGs) have increasing methylation rates as individuals age. We used elastic net models to select these sites among the thousands of CpGs sequenced across the full-genome data we have for our test individuals. 

## Cleaning Data Set
We've removed regions of the genome related to sex-determination, a putative inversion region, and mitochondrial and non-mapped regions to eliminate sources of bias within the genome (including CpGs that are related to sex-determination or inversion rather than age). Additionally, elastic net models do not accommodate missing data, so any CpG sites that are not represented in all sequenced individuals were removed.

```{r}
#remove CpGs with NAs for the model
nas <- colSums(is.na(allCpGs))
allCpGs1 <- allCpGs[,which(nas < 1)]

#remove non-mapped scaffold and mitochondrial data from data set
allCpGs1 <- allCpGs1[,which(!(grepl(pattern = "NW",
                                    colnames(allCpGs1),perl = T)))]
allCpGs1 <- allCpGs1[,which(!(grepl(pattern = "NC_009709.1",
                                    colnames(allCpGs1),perl = T)))]

# remove sex-det and inversion chromosomes
allCpGs1 <- allCpGs1[,which(!(grepl(pattern = "NC_047158.1",
                                    colnames(allCpGs1),perl = T)))]
allCpGs1 <- allCpGs1[,which(!(grepl(pattern = "NC_047162.1",
                                    colnames(allCpGs1),perl = T)))]

allCpGs1 <- allCpGs1 %>% dplyr::select(-filename)
```

## Prepping to Run Models
We log age data prior to running the model, and the CpG selected is repeated 100 times with different training and testing sets. This was repeated with sex-separated clocks as well (glmnet R script) but it is shorted to just the full model here since that is what we ended up using.
```{r}
#log age data
sexdata1 <- sexdata %>% 
  filter(Age <= 20 & DataType == "Wild") %>% 
  mutate(LogAge = log(Age),LogLen = log(Length))

# run with just ages 16-29, subset to Methyl 2025 samples
CpGsMeta <- sexdata1 %>% left_join(allCpGs1) %>% filter(DataType == "Wild" & Age <= 20)
```

```{r}
#create blank objects for storing outputs in loop
modelout <- data.frame(matrix(data = NA,nrow = 1000, ncol = 5))
colnames(modelout) <- c("rep","Train_cor","Test_cor","Train_MAE","Test_MAE","lambda.1se")
CpGlist <- list()

predout <- data.frame(indivname = sexdata1$indivname,
                      LogAge = sexdata1$LogAge)
```

## Run 100 Elastic Net Model Replicates
```{r}
##Loop for glmnet testing with all CpG sites ####
i <- 1
for (i in 1:1000) {
  print(i)
  #set a different seed so different train/test and a separate model are constructed
  set.seed(Sys.time())
  #randomly select individuals for testing
  testpick <- CpGsMeta %>% 
    group_by(Age,ObsSex) %>% 
    slice_sample(n = 1) %>% 
    ungroup()
  
  #sort other individuals into training data
  trainpick <- CpGsMeta[which(!(CpGsMeta$indivname %in% testpick$indivname)),]
  
  #select data and run glmnet to select sites
  y <- as.matrix(trainpick$LogAge)
  x <- as.matrix(dplyr::select(trainpick,starts_with("NC")))
  
  model1 <- cv.glmnet(x,y,family = "gaussian",alpha = 0.5,nfolds = 10,type.measure = "mae")
  coef.allCpG <- as.matrix(coef(model1,s = "lambda.1se"))
  
  bestCpGnames <- names(coef.allCpG[which(coef.allCpG > 0),])
  
  #predict testdat
  ytest <- testpick$LogAge
  xtest <- as.matrix(dplyr::select(testpick,starts_with("NC")))
  
  #predict and correlate train data
  pred_logage <- predict(model1,newx = x,type = "response",s = model1$lambda.1se,type.measure = "mae")
  train_cor <- cor.test(y,pred_logage)
  
  #predict and correlate test data
  pred_logage_test <- predict(model1,newx = xtest,type = "response",s = model1$lambda.min,type.measure = "mae")
  test_cor <- cor.test(ytest,pred_logage_test)
  
  #uncertainty for train and test
  train_MAE <- median(abs(exp(y) - exp(pred_logage[ ,1])))
  test_MAE <- median(abs(exp(ytest) - exp(pred_logage_test[ ,1])))
  
  ##save outputs
  modelout[i,5] <- model1$lambda.min
  CpGlist[[i]] <- bestCpGnames
  
  modelout[i,1] <- train_cor$estimate
  modelout[i,2] <- test_cor$estimate
  
  modelout[i,3] <- train_MAE
  modelout[i,4] <- test_MAE
  
  #save predictions
  #full model
  trainpred <- data.frame(indivname = trainpick$indivname,
                          LogAge = trainpick$LogAge,
                          predage = pred_logage[1:length(pred_logage)])
  testpred <- data.frame(indivname = testpick$indivname,
                          LogAge = testpick$LogAge,
                          predage = pred_logage_test[1:length(pred_logage_test)])
  predout_tmp <- rbind(trainpred,testpred)
  predout <- full_join(predout,predout_tmp, by = c("indivname","LogAge"))
}

hist(modelout$Train_cor)
hist(modelout$Test_cor)

#concatenate all the best CpGs
CpGlist_best <- c()
for (i in 1:1000) {
  tmpCpG <- CpGlist[[i]]
  names(tmpCpG) <- rep(i,times = length(tmpCpG))
  CpGlist_best <- c(CpGlist_best,tmpCpG)
}

counting <- data.frame(CpGlist_best) %>% 
  group_by_all() %>% 
  count() %>% 
  arrange(desc(n))
table(counting$n)

top_CpG_group <- counting %>% 
  arrange(desc(n)) %>% 
  filter(n > 100)

best_corrs <- modelout %>% 
  mutate(rep=seq(from = 1, to = nrow(modelout)),ttdiff = abs(Train_MAE-Test_MAE)) %>% 
  arrange(ttdiff) %>% 
  slice_head(n=20)

bestCpG = unique(CpGlist_best[which(names(CpGlist_best) %in% best_corrs$rep)])

best_corrsCpG <- data.frame(CpGlist_best=bestCpG,type="Age")
top_CpG_group$CpGlist_best %in% best_corrsCpG$CpGlist_best
best_corrsCpG$CpGlist_best %in% top_CpG_group$CpGlist_best

#!# keep best_corrs list
```

## Repeat Loop with Length
```{r}
#create blank objects for storing outputs in loop
modelout <- data.frame(matrix(data = NA,nrow = 1000, ncol = 6))
colnames(modelout) <- c("rep","Train_cor","Test_cor","Train_MAE","Test_MAE","lambda.1se")
CpGlist <- list()

predout <- data.frame(indivname = sexdata1$indivname,
                      LogLen = sexdata1$LogLen)

##Loop for glmnet testing with all CpG sites ####
i <- 1
for (i in 1:1000) {
  #print(i)
  set.seed(Sys.time())
  
  #randomly select individuals for testing
  testpick <- CpGsMeta %>% 
    group_by(Age,ObsSex) %>% 
    slice_sample(n = 1) %>% 
    ungroup()
  
  #sort other individuals into training data
  trainpick <- CpGsMeta[which(!(CpGsMeta$indivname %in% testpick$indivname)),]
  
  #select data and run glmnet to select sites
  y <- trainpick$LogLen
  x <- as.matrix(dplyr::select(trainpick,starts_with("NC")))
  
  model1 <- cv.glmnet(x,y,family = "gaussian",alpha = 0.5,nfolds = 10,type.measure = "mae")
  coef.allCpG <- as.matrix(coef(model1,s = "lambda.1se"))
  
  bestCpGnames <- names(coef.allCpG[which(coef.allCpG > 0),])
  
  #predict testdat
  ytest <- testpick$LogLen
  xtest <- as.matrix(dplyr::select(testpick,starts_with("NC")))
  
  #predict and correlate train data
  pred_LogLen <- predict(model1,newx = x,type = "response",s = model1$lambda.1se,type.measure = "mae")
  train_cor <- cor.test(y,pred_LogLen)
  
  #predict and correlate test data
  pred_LogLen_test <- predict(model1,newx = xtest,type = "response",s = model1$lambda.1se,type.measure = "mae")
  test_cor <- cor.test(ytest,pred_LogLen_test)
  
  #uncertainty for train and test
  train_MAE <- median(abs(exp(y) - exp(pred_LogLen[ ,1])))
  test_MAE <- median(abs(exp(ytest) - exp(pred_LogLen_test[ ,1])))
  
  ##save outputs
  modelout[i,6] <- model1$lambda.min
  modelout[i,1] <- i
  CpGlist[[i]] <- bestCpGnames
  
  modelout[i,2] <- train_cor$estimate
  modelout[i,3] <- test_cor$estimate
  
  modelout[i,4] <- train_MAE
  modelout[i,5] <- test_MAE
  
  #save predictions
  #full model
  trainpred <- data.frame(indivname = trainpick$indivname,
                          LogLen = trainpick$LogLen,
                          predlen = pred_LogLen)
  testpred <- data.frame(indivname = testpick$indivname,
                         LogLen = testpick$LogLen,
                         predlen = pred_LogLen_test)
  predout_tmp <- rbind(trainpred,testpred)
  predout <- full_join(predout,predout_tmp, by = c("indivname","LogLen"))
}

hist(modelout$Train_cor)
hist(modelout$Test_cor)

#concatenate all the best CpGs
CpGlist_best_len <- c()
for (i in 1:1000) {
  tmpCpG <- CpGlist[[i]]
  names(tmpCpG) <- rep(i,times = length(tmpCpG))
  CpGlist_best_len <- c(CpGlist_best_len,tmpCpG)
}

counting <- data.frame(CpGlist_best_len) %>% 
  group_by_all() %>% 
  count() %>% 
  arrange(desc(n))
table(counting$n)

top_CpG_group <- counting %>% 
  arrange(desc(n)) %>% 
  filter(n > 100)

best_corrs_len <- modelout %>% 
  mutate(rep=seq(from = 1, to = nrow(modelout)),ttdiff = abs(Train_MAE-Test_MAE)) %>% 
  arrange(ttdiff) %>% 
  slice_head(n=20)

bestCpG_len = unique(CpGlist_best_len[which(names(CpGlist_best_len) %in% best_corrs_len$rep)])

best_corrsCpG_len <- data.frame(CpGlist_best=bestCpG_len,type="Length")
top_CpG_group$CpGlist_best_len %in% best_corrsCpG_len$CpGlist_best_len
best_corrsCpG_len$CpGlist_best_len %in% top_CpG_group$CpGlist_best_len

#!# keep best_corrs list
```

## Concatenate the best CpGs across replicates
```{r}
CpGlist_best_comb <- unique(rbind(best_corrsCpG,best_corrsCpG_len))

write.csv(CpGlist_best_comb,file = "Output/CpGList_LengthAge_MAEBestModels.csv",row.names = F)
```

## Figures

### Correlation and MAE values for age prediction
```{r}
predout %>% 
  gather(key = "rep",value = "PredAge",-indivname,-LogLen) %>% 
  ggplot(aes(group = factor(exp(LogLen)),y = exp(PredAge),x=exp(LogLen))) +
    geom_boxplot() +
    #geom_jitter(aes(color=indivname)) +
    ggtitle("Halibut - Training data") +
    xlab("Otolith Age Estimate (years)") +
    ylab("Molecular clock Age (years)") +
    theme_classic(base_size = 18)

## quick look at lambda values, correlation levels and MAE values across replicates
modelout[,2:6] %>% 
  gather(key="stat",value = "value") %>% 
  ggplot(aes(x=stat,y=value))+
  geom_boxplot()+
  theme_classic(base_size = 18)
  
  
```


### Plotting individuals and percent methylation values for each CpG sites

```{r}
sexdata <- sexdata %>% 
  mutate(LogAge = log(Age),LogLen = log(Length))

bestCpGs <- allCpGs1[,which(colnames(allCpGs1) %in% CpGlist_best_comb$CpGlist_best)]
bestCpGs <- cbind(indivname=allCpGs1$indivname,bestCpGs)
tmp <- sexdata[,1:3]
jpeg(file = "Figures/PerCpGOtolith.jpeg", width = 1200, height = 1000)
tmp %>% full_join(bestCpGs,by = "indivname") %>% 
  gather(key = "CpG", value = "Methyl",-indivname,-ObsSex,-Age) %>% 
  #filter(CpG == "NC_047159.1.29166055" | CpG == "NC_047153.1.32666785" | CpG == "NC_047163.1.29070020") %>% 
  ggplot(aes(x = Age, y = Methyl, color = ObsSex))+
  geom_point()+
  facet_wrap(~CpG,scales = "free_y")+
  xlab("Otolith Age Estimate") +
  ylab("Percent Methylation") +
  ggtitle("Otolith Age vs. Percent Methylation per CpG") +
  theme_bw()+
  theme(legend.position = "none")
dev.off()

#similar plot for CpGs selected with just age 4 to 15 data
bestCpGs <- allCpGs1[,which(colnames(allCpGs1) %in% CpGlist_best_comb_age4to15$CpGlist_best)]
bestCpGs <- cbind(indivname=allCpGs1$indivname,bestCpGs)
tmp <- sexdata[,1:3]
jpeg(file = "Figures/PerCpGOtolith_Age4to15CpGs.jpeg", width = 1200, height = 1000)
tmp %>% full_join(bestCpGs,by = "indivname") %>% 
  gather(key = "CpG", value = "Methyl",-indivname,-ObsSex,-Age) %>% 
  #filter(CpG == "NC_047159.1.29166055" | CpG == "NC_047153.1.32666785" | CpG == "NC_047163.1.29070020") %>% 
  ggplot(aes(x = Age, y = Methyl, color = ObsSex))+
  geom_point()+
  facet_wrap(~CpG,scales = "free_y")+
  xlab("Otolith Age Estimate") +
  ylab("Percent Methylation") +
  ggtitle("Otolith Age vs. Percent Methylation per CpG") +
  theme_bw()+
  theme(legend.position = "none")
dev.off()
```

