---
title: "2. Age Prediction with Selected CpG Data"
author: "Ellie Weise"
date: "2024-09-20"
output: html_document
---

Once we run the elastic net model with all potential CpGs, we do the actual age prediction with those sites. Prediction can also be done with one-step, using the elastic net with all sites. But generally literature with whole-genome data has done two sets of elastic nets because the correlations are stronger with less noise, and the clock is able to be scaled without doing more whole genome sequencing. That is, we could do targeted sequencing and add it to the same clock instead of needing all the sites.

## Run Clock with Only Selected Sites

```{r,message=F,warning=FALSE}
#load libraries
library(tidyverse)
library(glmnet)
library(MASS)
library(wesanderson)
library(ggforce)
###run clock with best CpG sites
###run clock with best CpG sites
CpGlist_best_old <- read.csv(file = "Output/CpGList_LengthAge.csv")
CpGlist_best_old$Set <- "OldSites"
CpGlist_best_old <- CpGlist_best_old %>% dplyr::select(CpGlist_best,n,Set)
CpGlist_best_4to15 <- read.csv(file = "Output/CpGList_LengthAge_Age4to15_121025.csv")
CpGlist_best_4to15$Set <- "Age4to15"
CpGlist_best_4to29 <- read.csv(file = "Output/CpGList_LengthAge_Age4to29_121025.csv")
CpGlist_best_4to29$Set <- "Age4to29"
CpGlist_all <- rbind(CpGlist_best_4to15,CpGlist_best_4to29)
#CpGlist_all <- CpGlist_best_old
##load data
# if doing clock from scratch
load("MethylKitInput/allCpGs_MinimalMissing.rda")
sexdata <- read.csv("Input/MethylWild_Sex_AgeData.csv")

```

### Preparing data for elastic net
```{r}
## Data prep ####
set.seed(1230)
# subset to age 4-15 wild data
sexdata <- sexdata %>% 
  filter(DataType == "Wild" & Age < 20)

#running a glmnet with all selected sites
#CpGlist_all <- CpGlist_all %>% filter(ModelType == "Age")
bestCpGs <- allCpGs1[,which(colnames(allCpGs1) %in% CpGlist_all$CpGlist_best)]

#remove CpGs with NAs for the model
nas <- colSums(is.na(bestCpGs))
bestCpGs <- bestCpGs[,which(nas < 1)]
bestCpGs <- cbind(indivname=allCpGs1$indivname,bestCpGs)

#write.csv(allCpGs1,"allCpGs_nonas.csv",append = F,quote = F)

#!# remove sex-det and inversion!!!
#allCpGs1 <- allCpGs1[,which(!(grepl(pattern = "NC_047158.1",
#                                    colnames(allCpGs1),perl = T)))]
#allCpGs1 <- allCpGs1[,which(!(grepl(pattern = "NC_047162.1",
#                                    colnames(allCpGs1),perl = T)))]
# remove positive control from HPEI set
#allCpGs1 <- allCpGs1[-125,]

# check nas across individuals
# should be zero!!
#nas_row <- rowSums(is.na(allCpGs1))

#log age data
sexdata <- sexdata %>% 
  mutate(LogAge = log(Age),LogLen = log(Length))

sexdata_f <- sexdata %>% 
  filter(ObsSex == 2)

sexdata_m <- sexdata %>% 
  filter(ObsSex == 1)
```


### Separating into Training and Testing Data Sets 
```{r}
#randomly select individuals for testing
testpick <- sexdata %>% 
  group_by(Age,ObsSex) %>% 
  slice_sample(n = 1)

#sort other individuals into training data
trainpick <- sexdata[which(!(sexdata$indivname %in% testpick$indivname)),]

#select CpG sites for each group
testdat <- bestCpGs[which(bestCpGs$indivname %in% testpick$indivname),]
testdat <- left_join(testpick,testdat,by = "indivname")

traindat <- bestCpGs[which(!(bestCpGs$indivname %in% testpick$indivname)),]
traindat <- left_join(trainpick,traindat,by = "indivname")
```


### Run Elastic Net Model on Training Set and Predict Test Set
```{r,warning=F}
## Full Data Model ####
#select data and run glmnet to select sites
y <- traindat$LogAge

#running a glmnet with all selected sites
bestCpG_train <- as.matrix(traindat[,which(colnames(traindat) %in% CpGlist_all$CpGlist_best)])

table(colSums(is.na(bestCpG_train)))
nas <- colSums(is.na(bestCpG_train))
bestCpG_train <- bestCpG_train[,which(nas < 1)]

model1 <- cv.glmnet(bestCpG_train,y,family = "gaussian",alpha = 0.05,nfolds = 20)
coef.allCpG <- as.matrix(coef(model1,s = "lambda.min"))

bestCpGnames <- names(coef.allCpG[which(coef.allCpG > 0),])

#predict testdat
ytest <- testdat$LogAge
xtest <- as.matrix(testdat[,which(colnames(testdat) %in% CpGlist_all$CpGlist_best)])
nas <- colSums(is.na(bestCpG_train))
xtest <- xtest[,which(nas < 1)]

#predict and correlate train data
pred_logage <- predict(model1,newx = bestCpG_train,type = "response",s = model1$lambda.1se)
colnames(pred_logage) <- "Epi_LogAge"
train_cor <- cor.test(y,pred_logage)
plot(y,pred_logage,col = traindat$ObsSex)

plot(model1, xvar = "dev", label = TRUE)

#predict and correlate test data
pred_logage_test <- predict(model1,newx = xtest,type = "response",s = model1$lambda.1se)
colnames(pred_logage_test) <- "Epi_LogAge"
test_cor <- cor.test(ytest,pred_logage_test)
plot(ytest,pred_logage_test,col = testdat$ObsSex)

#uncertainty for train and test
train_MAE <- median(abs(exp(y) - exp(pred_logage[ ,1])))
test_MAE <- median(abs(exp(ytest) - exp(pred_logage_test[ ,1])))

```

### Repeat Clock for female and male-only data sets
```{r,warning=F}
##Sex-separated clocks ####
#female clock
trainf <- traindat %>% filter(ObsSex == 2)
testf <- testdat %>% filter(ObsSex == 2)
xf <- as.matrix(trainf[,which(colnames(trainf) %in% CpGlist_all$CpGlist_best)])
yf <- trainf$LogAge

ytestf <- testf$LogAge
xtestf <- as.matrix(testf[,which(colnames(testf) %in% CpGlist_all$CpGlist_best)])

modelf <- cv.glmnet(xf,yf,family = "gaussian",alpha = 0.05,nfolds = 10)

#predict train and test correlation
pred_logage_f <- predict(modelf,newx = xf,type = "response",s = model1$lambda.1se)
colnames(pred_logage_f) <- "Epi_LogAge"
train_cor_f <- cor.test(yf,pred_logage_f)
#plot(yf,pred_logage_f)

#predict test and correlation
pred_logage_test_f <- predict(modelf,newx = xtestf,type = "response",s = model1$lambda.1se)
colnames(pred_logage_test_f) <- "Epi_LogAge"
test_cor_f <- cor.test(ytestf,pred_logage_test_f)
#plot(ytestf,pred_logage_test_f)

#uncertainty for train and test
train_MAE_f <- median(abs(exp(yf) - exp(pred_logage_f[ ,1])))
test_MAE_f <- median(abs(exp(ytestf) - exp(pred_logage_test_f[ ,1])))


#male clock
trainm <- traindat %>% filter(ObsSex == 1)
testm <- testdat %>% filter(ObsSex == 1)
xm <- as.matrix(trainm[,which(colnames(trainm) %in% CpGlist_all$CpGlist_best)])
ym <- trainm$LogAge

ytestm <- testm$LogAge
xtestm <- as.matrix(testm[,which(colnames(testm) %in% CpGlist_all$CpGlist_best)])

#run glmnet model with alpha=0 to include all sites
modelm <- cv.glmnet(xm,ym,family = "gaussian",alpha = 0.05,nfolds = 10)

#predict train and correlate
pred_logage_m <- predict(modelm,newx = xm,type = "response",s = model1$lambda.1se)
colnames(pred_logage_m) <- "Epi_LogAge"
train_cor_m <- cor.test(ym,pred_logage_m)$estimate
#plot(ym,pred_logage_m)

#predict test and correlate
pred_logage_test_m <- predict(modelm,newx = xtestm,type = "response",s = model1$lambda.1se)
colnames(pred_logage_test_m) <- "Epi_LogAge"
test_cor_m <- cor.test(ytestm,pred_logage_test_m)
#plot(ytestm,pred_logage_test_m)

#uncertainty for clock
train_MAE_m <- median(abs(exp(ym) - exp(pred_logage_m[ ,1])))
test_MAE_m <- median(abs(exp(ytestm) - exp(pred_logage_test_m[ ,1])))

```

## Figures
### correlation and MAE values for all three models
```{r}
##Plots ####
## make results into dataframes
#clock correlation plot
train_res <- data.frame(indivname=traindat$indivname,
                        ObsSex=traindat$ObsSex,
                        Length=traindat$Length,
                        LogAge=y,
                        Epi_LogAge=pred_logage,
                        type="Train",
                        sampletype=traindat$DataType)
test_res <- data.frame(indivname=testdat$indivname,
                       ObsSex=testdat$ObsSex,
                       Length=testdat$Length,
                       LogAge=ytest,
                       Epi_LogAge=pred_logage_test,
                       type="Test",
                       sampletype=testdat$DataType)
res_comb_all <- rbind(train_res,test_res)
res_comb_all <- res_comb_all %>% mutate(Age=exp(LogAge),EpiAge=exp(Epi_LogAge))
res_comb_all$ModelName <- "Combined"

train_res_f <- data.frame(indivname=trainf$indivname,
                          ObsSex=trainf$ObsSex,
                          Length=trainf$Length,
                          LogAge=yf,
                          Epi_LogAge=pred_logage_f,
                          type="Train", 
                          sampletype=trainf$DataType)
test_res_f <- data.frame(indivname=testf$indivname,
                         ObsSex=testf$ObsSex,
                         Length=testf$Length,
                         LogAge=ytestf,
                         Epi_LogAge=pred_logage_test_f,
                         type="Test",
                         sampletype=testf$DataType)
res_comb_all_f <- rbind(train_res_f,test_res_f)
res_comb_all_f <- res_comb_all_f %>% mutate(Age=exp(LogAge),EpiAge=exp(Epi_LogAge))
res_comb_all_f$ModelName <- "Female Only"

train_res_m <- data.frame(indivname=trainm$indivname,
                          ObsSex=trainm$ObsSex,
                          Length=trainm$Length,
                          LogAge=ym,
                          Epi_LogAge=pred_logage_m,
                          type="Train",
                          sampletype=trainm$DataType)
test_res_m <- data.frame(indivname=testm$indivname,
                         ObsSex=testm$ObsSex,
                         Length=testm$Length,
                         LogAge=ytestm,
                         Epi_LogAge=pred_logage_test_m,
                         type="Test",
                         sampletype=testm$DataType)
res_comb_all_m <- rbind(train_res_m,test_res_m)
res_comb_all_m <- res_comb_all_m %>% mutate(Age=exp(LogAge),EpiAge=exp(Epi_LogAge))
res_comb_all_m$ModelName <- "Male Only"

res_merged <- rbind(res_comb_all,res_comb_all_f,res_comb_all_m)
#jpeg(file = "Figures/AgingClock_Combined_AgeLengthCpGs_NewRun_AddHatchery_Age4to19_Age4to15CpGs.jpeg", width = 900, height = 480)
ggplot(res_merged,aes(x = Age,y = EpiAge,color=as.character(ObsSex),shape = type))+
  facet_grid(~ModelName) +
  geom_point(size = 3.5)+
  geom_jitter(height = 0,width = 0.1)+
  ylim(0, 30) + xlim(0, 30) +
  geom_abline(slope = 1,intercept = 0) +
  xlab("Otolith Age (years)") +
  ylab("Molecular clock Age (years)") +
  ggtitle("Correlation Between Otolith and Molecular Clock Age") +
  scale_color_manual(values = c(wes_palette("IsleofDogs1")[1],wes_palette("IsleofDogs1")[3]))+
  theme_classic(base_size = 18) + 
  theme(legend.position = "none")

#dev.off()

#MAE values
res_merged$AbsErr <- abs(res_merged$Age - res_merged$EpiAge)
res_merged$indivname[res_merged$AbsErr > 5]
#jpeg(file = "Figures/MAEPlot_AgeLengthCpGs_NewRun_AddHatchery_Age4to19_Age4to15CpGs.jpeg", width = 900, height = 480)

ggplot(res_merged,aes(x = type, y = AbsErr)) +
  geom_boxplot() +
  geom_jitter(aes(color = as.character(ObsSex)), size = 3) +
  facet_wrap(~ModelName) +
  scale_color_manual(values = c(wes_palette("IsleofDogs1")[1],wes_palette("IsleofDogs1")[3])) +
  theme_classic(base_size = 18) + 
  labs(color = "Sex", x = element_blank(), y = "Absolute Error")
#dev.off()

c(train_MAE,test_MAE,train_MAE_f,test_MAE_f,train_MAE_m,test_MAE_m)
```

```

