---
title: "4. Analyzing Nanopore Run Data"
author: "Ellie Weise"
date: "2024-09-20"
output: html_document
---

## Overview
Once Nanopore data has been processed into an individual vs. percent methylation table, we can look at both that and the per adaptive region depth value to evaluate the success of the nanopore run.

## Testing Nanopore percent methylation against existing clock

### loading in data and libraries
```{r}
### Testing Nanopore results with existing model
#load packages
library(tidyverse)
library(glmnet)
library(MASS)
library(paletteer)

#load data
load("Output/AgeLengthCpG_modelrun_021524.RData")
barcode1 <- read.table("Input/Barcode3_21-SS-HIP-5533_v5.bedMethyl",header = F,stringsAsFactors = F)
barcode1 <- subset(barcode1,barcode1$V4 == "h")
barcode2 <- read.table("Input/Barcode4_20-SS-HS-HYP157_v5.bedMethyl",header = F,stringsAsFactors = F)
barcode2 <- subset(barcode2,barcode2$V4 == "h")
#selecting just percent mod and CpG position
pm1 <- barcode1[,c(1,3,11)]
pm2 <- barcode2[,c(1,3,11)]
colnames(pm1) <- c("CHROM","END","PERCMOD")
colnames(pm2) <- c("CHROM","END","PERCMOD")
```

### Reformatting to match the Illumina data
```{r}
#reformatting to match Illumina individual data
pm1 <- pm1 %>% 
  mutate(CpG = paste(CHROM,END,sep = ".")) %>% 
  dplyr::select(CpG,PERCMOD) %>% 
  spread(key = CpG,value = PERCMOD)

pm2 <- pm2 %>% 
  mutate(CpG = paste(CHROM,END,sep = ".")) %>% 
  dplyr::select(CpG,PERCMOD) %>% 
  spread(key = CpG,value = PERCMOD)

best1 <- pm1[1,which(colnames(pm1) %in% CpGlist_best_comb$CpGlist_best)]
best2 <- pm2[1,which(colnames(pm2) %in% CpGlist_best_comb$CpGlist_best)]
```

### Running elastic net model with shared sites and predicting nanopore ages
```{r}
#select data and run glmnet to select sites
y <- traindat$LogAge

#running a glmnet with all selected sites
bestCpG_train <- as.matrix(traindat[,which(colnames(traindat) %in% names(best1))])
model1 <- cv.glmnet(bestCpG_train,y,family = "gaussian",alpha = 0.05,nfolds = 10)
coef.allCpG <- as.matrix(coef(model1,s = "lambda.min"))

bestCpGnames <- names(coef.allCpG[which(coef.allCpG > 0),])

#predict testdat
ytest <- c(testdat$LogAge,log(4),log(15))
xtest <- as.matrix(testdat[,which(colnames(testdat) %in% names(best1))])
xtest <- rbind(xtest,as.numeric(best1),as.numeric(best2))

#predict and correlate train data
pred_logage <- predict(model1,newx = bestCpG_train,type = "response",s = model1$lambda.min)
train_cor <- cor.test(y,pred_logage)
#plot(y,pred_logage,col = traindat$ObsSex)

plot(model1, xvar = "dev", label = TRUE)

#predict and correlate test data
pred_logage_test <- predict(model1,newx = xtest,type = "response",s = model1$lambda.min)
test_cor <- cor.test(ytest,pred_logage_test)

plot(ytest,pred_logage_test)
```

*We found that the Nanopore percent methylation values were correlated with the Illumina data for both our individuals, but the percent methylation values themselves are different enough that they can't be combined into a single clock. Other groups have had similar difficulties combining methylation data across platforms. However, it has been shown that the results and CpG locations can be corroborated across platforms, just not the percent methylation values themselves.*

## Testing per CpG depth for Nanopore Runs
### Reading in depth files and adaptive files
```{r}
depth1 <- read.table("Input/Barcode1_21-SS-HIP-5533.txt")
colnames(depth1) <- c("CHROM","POS","DEPTH")

depth2 <- read.table("Input/Barcode2_20-SS-HS-HYP157.txt")
colnames(depth2) <- c("CHROM","POS","DEPTH")

depth3 <- read.table("Input/Barcode3_21-SS-HIP-5533_v5_depth.txt")
colnames(depth3) <- c("CHROM","POS","DEPTH")

depth4 <- read.table("Input/Barcode4_20-SS-HS-HYP157_v5_depth.txt")
colnames(depth4) <- c("CHROM","POS","DEPTH")

adapt1 <- read.table("Output/NanoporeRegionSelection_040424.txt")
colnames(adapt1) <- c("CHROM", "START", "END")

adapt2 <- read.table("Output/NanoporeRegionSelection_061224.txt")
colnames(adapt2) <- c("CHROM", "START", "END")
```

### match depth to nanopore adaptive regions
```{r}
## match file to adaptive sequencing
adapt1$LENGTH <- adapt1$END - adapt1$START
adapt1$NAME <- paste0(adapt1$CHROM,":",adapt1$START,"-",adapt1$END,",", adapt1$LENGTH)
adapt2$LENGTH <- adapt2$END - adapt2$START
adapt2$NAME <- paste0(adapt2$CHROM,":",adapt2$START,"-",adapt2$END,",", adapt2$LENGTH)

depth1$NAME <- NA
depth2$NAME <- NA
depth3$NAME <- NA
depth4$NAME <- NA

for (i in 1:nrow(adapt1)) {
  depth1$NAME[which(depth1$CHROM == adapt1$CHROM[i] &
                      depth1$POS >= adapt1$START[i] &
                      depth1$POS < adapt1$END[i])] <- adapt1$NAME[i]
  
  depth2$NAME[which(depth2$CHROM == adapt1$CHROM[i] &
                      depth2$POS >= adapt1$START[i] &
                      depth2$POS < adapt1$END[i])] <- adapt1$NAME[i]
}

for (i in 1:nrow(adapt2)) {
  depth3$NAME[which(depth3$CHROM == adapt2$CHROM[i] &
                      depth3$POS >= adapt2$START[i] &
                      depth3$POS < adapt2$END[i])] <- adapt2$NAME[i]
  
  depth4$NAME[which(depth4$CHROM == adapt2$CHROM[i] &
                      depth4$POS >= adapt2$START[i] &
                      depth4$POS < adapt2$END[i])] <- adapt2$NAME[i]
}
```

### Put files together
```{r}
summ1 <- depth1 %>% 
  group_by(NAME) %>% 
  summarise(avg1 = mean(DEPTH),
            med1 = median(DEPTH)) %>% 
  separate(col = NAME,into = c("NAME","LENGTH"),sep = ",")

summ2 <- depth2 %>% 
  group_by(NAME) %>% 
  summarise(avg2 = mean(DEPTH),
            med2 = median(DEPTH)) %>% 
  separate(col = NAME,into = c("NAME","LENGTH"),sep = ",")

summ3 <- depth3 %>% 
  group_by(NAME) %>% 
  summarise(avg3 = mean(DEPTH),
            med3 = median(DEPTH)) %>% 
  separate(col = NAME,into = c("NAME","LENGTH"),sep = ",") %>% 
  filter(!is.na(NAME))

summ4 <- depth4 %>% 
  group_by(NAME) %>% 
  summarise(avg4 = mean(DEPTH),
            med4 = median(DEPTH)) %>% 
  separate(col = NAME,into = c("NAME","LENGTH"),sep = ",")%>% 
  filter(!is.na(NAME))



summall <- full_join(summ1,summ2,by = c("NAME","LENGTH"))
summall <- full_join(summall,summ3,by = c("NAME","LENGTH"))
summall <- full_join(summall,summ4,by = c("NAME","LENGTH"))

summ1$Barcode <- "Barcode1"
summ2$Barcode <- "Barcode2"
summ3$Barcode <- "Barcode3"
summ4$Barcode <- "Barcode4"

depth3$Barcode <- "Barcode3"
depth4$Barcode <- "Barcode4"
```


### Figures
```{r}
rbind(depth3,depth4) |>
  separate(col = NAME,into = c("NAME","LENGTH"),sep = ",") %>% 
  group_by(NAME,Barcode) %>% 
  reframe(spot=POS-min(POS)+1,
            DEPTH=DEPTH) %>% 
  filter(!is.na(NAME)) %>% 
  ggplot(aes(x=spot,y=DEPTH,color=Barcode))+
  facet_wrap(~NAME,scales = "free")+
  geom_line(alpha=0.6) +
  theme_bw()

rbind(depth1,depth2) |>
  separate(col = NAME,into = c("NAME","LENGTH"),sep = ",") %>% 
  group_by(NAME,Barcode) %>% 
  reframe(spot=POS-min(POS)+1,
          DEPTH=DEPTH,
          LENGTH=LENGTH) %>% 
  filter(!is.na(NAME)) %>% 
  ggplot(aes(x=spot,y=DEPTH,group=NAME,color=as.numeric(LENGTH)))+
  facet_wrap(~Barcode,scales = "free")+
  geom_line()+
  scale_color_viridis_c(option="magma")+
  theme_bw()

rbind(depth3,depth4) |>
  separate(col = NAME,into = c("NAME","LENGTH"),sep = ",") %>% 
  group_by(NAME,Barcode) %>% 
  reframe(spot=POS-min(POS)+1,
          DEPTH=DEPTH,
          LENGTH=LENGTH) %>% 
  filter(!is.na(NAME)) %>% 
  ggplot(aes(x=spot,y=DEPTH,group=NAME,color=as.numeric(LENGTH)))+
  facet_wrap(~Barcode,scales = "free")+
  geom_line()+
  scale_color_viridis_c(option="magma")+
  theme_bw()
 
##Pairwise plots
par(mfrow = c(2,3))
plot(summ1$med1,summ2$med2,xlab = "Barcode 1 - Age 4", ylab = "Barcode 2 - Age 15",xlim = c(0,600),ylim = c(0,600))
plot(summ1$med1,summ3$med3,xlab = "Barcode 1 - Age 4", ylab = "Barcode 3 - Age 4",xlim = c(0,600),ylim = c(0,600))
plot(summ1$med1,summ4$med4,xlab = "Barcode 1 - Age 4", ylab = "Barcode 4 - Age 15",xlim = c(0,600),ylim = c(0,600))
plot(summ2$med2,summ3$med3,xlab = "Barcode 2 - Age 15", ylab = "Barcode 3 - Age 4",xlim = c(0,600),ylim = c(0,600))
plot(summ2$med2,summ4$med4,xlab = "Barcode 2 - Age 15", ylab = "Barcode 4 - Age 15",xlim = c(0,600),ylim = c(0,600))
plot(summ3$med3,summ4$med4,xlab = "Barcode 3 - Age 4", ylab = "Barcode 4 - Age 15",xlim = c(0,600),ylim = c(0,600))
```

```{r}
depall <- rbind(depth1,depth2)
depall <- rbind(depall,depth3)
depall <- rbind(depall,depth4)

CpGs_clock <- read.csv("Output/CpGList_LengthAge.csv")

depth_CpGs <- depall[which(depall$CHROM == CpGs_clock$CHROM[1] & depall$POS == CpGs_clock$POS[1]),]

for (i in 2:nrow(CpGs_clock)) {
  depth_CpGs <- rbind(depth_CpGs,depall[which(depall$CHROM == CpGs_clock$CHROM[i] & depall$POS == CpGs_clock$POS[i]),])
}

depth_CpGs$type <- paste(depth_CpGs$RUN,depth_CpGs$INDIVID,sep = ": ")
depth_CpGs$CpG <- paste(depth_CpGs$CHROM,depth_CpGs$POS,sep = "_")

depth_CpGs$Depth <- NA

depth_CpGs$Depth[which(depth_CpGs$DEPTH < 50)] <- "0-50"
#depth_CpGs$Depth[which(depth_CpGs$DEPTH > 50 & depth_CpGs$DEPTH <= 100)] <- "50-100"
#depth_CpGs$Depth[which(depth_CpGs$DEPTH > 100 & depth_CpGs$DEPTH <= 200)] <- "100-200"
#depth_CpGs$Depth[which(depth_CpGs$DEPTH > 200 & depth_CpGs$DEPTH <= 300)] <- "200-300"
#depth_CpGs$Depth[which(depth_CpGs$DEPTH > 300 & depth_CpGs$DEPTH <= 400)] <- "300-400"
depth_CpGs$Depth[which(depth_CpGs$DEPTH >= 50 & depth_CpGs$DEPTH <= 500)] <- "50-500"
depth_CpGs$Depth[which(depth_CpGs$DEPTH > 500)] <- "500+"

#tiff(filename = "Figures/NanoporeDepthComp_justCpGs.tiff",width = 150,height = 150,units = "mm",res = 400)
ggplot(depth_CpGs,aes(x=type,y=CpG,fill = Depth))+
  geom_tile(color = "white",
            lwd = 0.5,
            linetype = 1)+
  facet_wrap(~type,scales = "free_x",ncol = 4)+
  #coord_fixed(expand = FALSE) +
  scale_fill_paletteer_d("nationalparkcolors::Acadia")+
  labs(y = "CpG site in Epigenetic Age Prediction (n = 87)")+
  scale_x_discrete(position = "top") +
  ggtitle("Nanopore Sequencing Depth at Targeted CpG Sites")+
  theme(axis.text.y = element_blank(),axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.y = element_text(margin = margin(r = -5), size = 12),
        strip.text = element_text(size = 12,color = "black"),
        strip.background = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(size = 14,face = "bold"),
        legend.position = "none")
#dev.off()
```

